\section{Code development}
\subsection{Development cycle}
Development of a new capability consists of several steps that are summarized below.
Some steps can be skipped during a casual work cycle of code support, bug fixes, and minor improvements. 

\begin{itemize}
\item Create a new github development branch.
\item Create github ticket or multiple tickets that summarize and stage the development process.
\item Implement numerical algorithms and add them to an Amanzi library.
\item Write unit tests for the new code.
\item Integrate new functionality into other algorithms.
\item Write integrated unit tests as needed.
\item If implemented algorithms take control parameters from an XML file, 
      document these parameters.
\item Test new capability and add a benchmark or verification test to the
      user guide.
\item Create a pull request to inform team members about the new capability and 
      to collect miscallenous feedback.
\item Merge the development branch into the master branch.
\end{itemize}


\subsection{Smart pointers}
We suggest the following guidelines on smart pointers and argument passing.

\begin{lstlisting}
unique_ptr<X> factory();      // creates an X
void sink(unique_ptr<X>&);    // consumes X, the caller cannot keep X 
                              // or use it after call
void reseat(unique_ptr<X>&);  // change which X the caller points to 

Teuchos::RCP<X> factory();  // creates an X
Teuchos::RCP<X> share();    // the caller will or might keep X
void share(const Teuchos::RCP<X>&);  // the callee will or might keep X
void reseat(Teuchos::RCP<X>&);       // change which X the caller points to 

void optional(X*);  // if X is an optional argument, i.e. pointer can be null

void if_primitive(X);  // if none of the above apply, and X is a primitive 
void if_not_primitive(X&);  // if none of the above apply, and X is not a primitive
X = primitive_return();       // if primitive
X& = non_primitive_return();  // if there is no ownership transfer implied,
                              // the caller just looks at (const) or modifies 
                              // (nonconst) X but does not keep it.
\end{lstlisting}


Note that X can be replaced by const X with no loss of generality.
Note that the key part of this is that passing by RCP should mean something about ownership, 
i.e. that the callee might keep X! 
This also suggests to NEVER use Teuchos::Ptr, instead using X*. 
Using Ptr requires that the object WAS stored in an RCP, and means an object stored in 
a unique\_ptr must jump through hoops to be put in a non-owning Ptr before it can be passed.


