\documentclass[12pt]{article}

% LaTeX Packages
\usepackage{color}
\usepackage[breaklinks]{hyperref}



% begin preamble
\setlength{\baselineskip}{16.0pt}    % 16 pt usual spacing between lines

\setlength{\parskip}{3pt plus 2pt}
\setlength{\parindent}{20pt}
\setlength{\oddsidemargin}{0.5cm}
\setlength{\evensidemargin}{0.5cm}
\setlength{\marginparsep}{0.75cm}
\setlength{\marginparwidth}{2.5cm}
\setlength{\marginparpush}{1.0cm}
\setlength{\textwidth}{150mm}

\pagestyle{empty} % use if page numbers not wanted

% end preamble

\begin{document}
\title{Amanzi Extern Software Requirements}

\maketitle


\section*{System Software Environment}
The libraries and tools listed in this section are assumed to be part of the target platform. Typically, 
a user would not build these libraries from source.  The Amanzi team recommends that users use
pre-built binaries or tools such as MacPorts, yum, rpm to install these libraries.  

At a minimum the build environment must have the following system software libraries and tools present.
\begin{itemize}
\item A UNIX-like OS. Amanzi has been built on several flavors of Linux (Red Hat/CentOS, Ubuntu, Fedora) and 
Mac OS 10.5 and 10.6. Native Windows builds are not supported at this time.
\item C and C++ compiler.
\item Python 2.6 version or higher.
\item MPI. Amanzi has been built against OpenMPI and MPICH. Amanzi does not support non-MPI builds at this time.
\item BLAS/LAPACK built and tuned to the target architecture.
\end{itemize}

If these libraries are installed in non-standard locations, user's will need to know the location 
of the header and library files to build most of the other third party libraries (TPLs).

\section*{Required External Software}
Amanzi leverages existing software libraries whenever possible. Amanzi can not built until all the required 
libraries are built successfully. In some cases (netCDF, Exodus, HDF5 and Trilinos), the libraries must be configured
in a particular way to work correctly with Amanzi and other TPLs. The Amanzi source directory contains a 
script that automatically builds the TPLs correctly. Users are strongly encouraged to use this tool to build the
TPLs. The current documentation for a correct TPL configuration is tracked in this tool. Advanced users creating
their own build process should refer to this script for supported configurations and TPL build order.

Amanzi builds the source located in \url{https://software.lanl.gov/ascem/trac/browser} every hour on several OS types.
The build and test suite results for each OS is recorded and tracked on the ASCEM Trac
site at \url{https://software.lanl.gov/ascem/trac/buildbot}. 

\begin{table}[htdp]
\begin{center}
\begin{tabular}{|c|c|p{3cm}|p{3cm}|p{3cm}|}
\hline
\hline
Name & Version & Description & Dependencies & Source Location \\
\hline
CMake & 2.8 & Cross platform build system & Python 2.6 & \url{http://www.cmake.org/cmake/resources/software.html}\\
\hline
Boost & 1.44.0 & Supplemental C++ Library & C++ &  \url{http://www.boost.org/users/download/}\\
\hline
zlib & 1.2.5 &Compression library (required by HDF5) & C, MPI \& zlib &  \url{http://zlib.net/} \\
\hline
cURL & 7.21.2 & Download wizard library (required netCDF) & C & \url{http://curl.haxx.se/download.html} \\
\hline
HDF5 &1.8.7 & File format library & C, MPI \& zlib & \url{http://www.hdfgroup.org/HDF5/release/obtain5.html} \\
\hline
netCDF & 4.1.1 & File format library & C, MPI, cURL \& HDF5 & \url{http://www.unidata.ucar.edu/software/netcdf/} \\
\hline
ExodusII & 4.98 & Mesh database library & C, MPI \& netCDF & \url{http://sourceforge.net/projects/exodusii/}\\
\hline
Trilnos & 10.6 & C++ numerical software library & C++, MPI, Exodus (if STKMesh is enabled) \& CMake & \url{http://trilinos.sandia.gov/download/trilinos-10.6.html} \\
\hline
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%


\section*{Supported External Software}
The TPLs listed in this section are not required to build Amanzi, but are required to support some Amanzi features. The 
library configuration for each of these is also listed in the Amanzi TPL build script. Please note the dependencies. Enabling
these TPLs in the build will likely require a change to the TPL build order. Please refer to the Amanzi TPL build script for
more information.

Although all the supported mesh framework libraries are supported in this section, users are required to build
Amanzi with at least one of the frameworks. 

\begin{table}[htdp]
\begin{center}
\begin{tabular}{|c|c|p{3cm}|p{3cm}|p{3cm}|}
\hline
\hline
Name & Version & Description & Dependencies & Source Location \\
\hline
UnitTest++ & 1.4 & C++ unit test framework & C++ & \url{http://sourceforge.net/projects/unittest-cpp/} \\
\hline
MOAB & rev 4225 (SVN) & Mesh framework & C/C++, MPI, Exodus & Tarfile located on Trac site \url{https://software.lanl.gov/ascem/trac/wiki/Amanzi/Building/TPL/MOAB}  \\
\hline
MSTK & 1.80 & Mesh framework & C, MPI, Exodus & \url{https://software.lanl.gov/MeshTools/trac}\\
\hline
STKMesh & & Mesh framework available in Trilinos & Exodus, Trilinos & Build with Trilinos with '-D ENABLE\_STKMesh:bool=true'  \\
\hline
SWIG &&&& \\
\hline
VisIt & 2.2.2 & visualization program & binaries available &  \url{https://wci.llnl.gov/codes/visit/home.html} \\
\hline
ASCEM-IO &&&& \\
\hline
DOxygen &&&& \\

&&&& \\
\hline


\end{tabular}
\end{center}
\label{default}
\end{table}%



\end{document}

